Manuscript: Detecting differences in Size Spectra

### 2023-07-18 Manuscript.tex
unstaged changes to manuscript before 1st submission

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
  \usepackage{amssymb}
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Manuscript: Detecting differences in Size Spectra},
  pdfauthor={Justin Pomeranz1,; James R. Junker2,3; Vojsava Gjoni4; Jeff S. Wesner4},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{lineno}
\usepackage{amsmath}
\usepackage{indentfirst}
\linenumbers
\newcommand{\beginsupplement}{ \setcounter{table}{0} \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\title{Manuscript: Detecting differences in Size Spectra}
\author{Justin Pomeranz\textsuperscript{1,*} \and James R.
Junker\textsuperscript{2,3} \and Vojsava
Gjoni\textsuperscript{4} \and Jeff S. Wesner\textsuperscript{4}}
\date{23 February, 2023}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The distribution of body size in communities is remarkably consistent
  across habitats and taxa and can be represented by size spectra, which
  are described by a power law. The focus of size spectra analysis is to
  estimate the exponent (\(\lambda\)) of the power law.
\item
  Many methods have been proposed for estimating \(\lambda\) most of
  which involve binning the data, summing abundance within bins, and
  then fitting a ordinary least squares (OLS) regression in log-log
  space. However, recent work has shown that binning procedures may
  return biased estimates of size spectra exponents compared to
  procedures that directly estimate \(\lambda\) using maximum likelihood
  estimation (MLE). Despite this variability in estimates, it is unclear
  if the relative change across environmental gradients is consistent
  across methodologies. Here, we used simulation to compare the ability
  of two binning methods (NAS, ELBn) and MLE to 1) recapture known
  values of \(\lambda\), and 2) recapture parameters in a linear
  regression measuring the change in \(\lambda\) across a hypothetical
  environmental gradient. We also compared the methods using two
  previously published body size datasets across a pollution gradient
  and a temperature gradient
\item
  Maximum likelihood methods always performed better than common binning
  methods, which demonstrated consistent bias depending on the simulated
  values of \(\lambda\). This bias carried over to the regressions,
  which were more accurate when \(\lambda\) was estimated using MLE
  compared to the binning procedures. Additionally, the variance in
  estimates using MLE methods is markedly reduced when compared to
  binning methods.
\item
  The uncertainty and variation in estimates when using binning methods
  is often greater than or equal to the variation previously published
  in experimental and observational studies, bringing into question the
  effect size of previously published results. However, while the
  methods produced different slope estimates from previously published
  datasets, they were in qualitative agreement on the sign of those
  slopes (i.e., all negative or all positive). Our results provide
  further support for the direct estimation of \(\lambda\) using MLE (or
  similar procedures) over the more common methods of binning.
\end{enumerate}

@@

We had Two primary objectives in this study: 1) to compare how well
different methods estimate site-specific \(\lambda\)'s 2) To compare how
different methods of estimating \(\lambda\) impact subsequent
regressions that use \(\lambda\)'s as a response variable. We did this
using both simulated and empirical data sets.

@@

After simulating data, we used three different methods (described below)
to estimate the value of \(\lambda\) (maximum likelihood, equal
logarithmic bins, and normalized abundance spectra). We repeated this
process 1000 times (reps) and plotted the distribution of values
obtained for each method.

@@

\hypertarget{equal-logarithmic-bins-elbn}{%
\subsubsection{Equal Logarithmic Bins:
ELBn}\label{equal-logarithmic-bins-elbn}}

In addition, to estimate the OLS slope parameter in log-log space, we
used two common binning methods. . It is important to note that the
log-log regression is estimating \(\lambda\) as the regressionslope
(\(\beta_{1}\)), and not \(\lambda\) directly. In the interest of
clarity, we will refer to all estimates of lambda as \(\lambda\)
regardless of the method used, and reserve the terms ``parameters,''
``intercept,'' and ``slope'' \(\beta_{env}\) to refer to the parameters
of linear regression that uses the lambdas as a response variable.

For the first binning method, we created 6 equal logarithmic bins
covering the range of body sizes. This method has been used extensively
in previous studies. For example see (Dossena et al. 2012, Martínez et
al. 2016, Perkins et al. 2018). In the present study, the count in each
bin was normalized by dividing by the bin width to account for the
unequal bin sizes. The process of normalizing shifts the estimate by -1.
In other words, an un-normalized estimate of -0.75 would result in an
estimate of -1.75 when normalizing the data (Sprules and Barth 2015,
Edwards et al. 2017, Pomeranz et al. 2022). Throughout the manuscript,
the normalized equal logarithmic binning method will be referred to as
ELBn.

@@

The second method was similar to ELBn, but bins of log\textsubscript{2}
widths are used. The number of bins is dependent on the range of body
sizes present in the data. When working with empirical data with
different size ranges this can alter the number of bins per site, and it
is recommended to construct the bins based on the global min and max
body size present. However, since the data here were simulated from a
known size range, the number of bins for each site is identical. The
count in each bin is normalized in the same way as described above. This
method has been used extensively in the literature e.g., (Jennings et
al. 2002, Jennings and Blanchard 2004, Sprules and Barth 2015, McGarvey
and Kirk 2018, Pomeranz et al. 2019a) and is referred to as the
Normalized Abundance Spectrum (NAS).

@@

Where \(N\) is the count in each bin, \(M\) is the mid-point of each
bin, and \(\beta_{1}\) is considered to be the \(\lambda_{estimate}\).

@@

Previous work has investigated the biases in \(\lambda\) estimates when
using different methods White et al. (2007). However, the focus of the
present work is to investigate biases when estimating the relationship
parameters across an environmental gradient.

@@

\[\lambda_{ik} = \beta_{0} + \beta_{0}x_{ik}\]

@@

Once the body size data were simulated, we then reversed the procedure.
First, we used the three methods (NAS, ELBn, MLE) to estimate
\(\lambda\) for each of the 5000 body size datasets. Then we fit a
linear regression to those data with the predictor variable \(x\) and
compared the resulting intercept (\(\beta_{0}\)) and slope
(\(\beta_{1}\)) parameter estimates to the known value. The main results
presented here were not dependent on the range of x-values or the number
of sites (Supplemental Information).

@@

The second data set was from the wadeable stream sites of NEON (National
Ecological Observatory Network (NEON) 2022). NEON stream sites are
located across a wide temperature gradient in the United States, from
Puerto Rico to Alaska. Quantitative macroinvertebrate samples were
collected using the most appropriate method based on the local habitat.
All individuals were identified and had their body lengths measured, and
body mass was estimated using published length weight regressions. This
data has been analyzed previously using size spectra methods as
described in (Pomeranz et al. 2022). Detailed methods of the sampling
collection and data processing methods can be in the macroinvertebrate
DPI pubs found on the NEON website
\url{https://www.neonscience.org/data-collection/macroinvertebrates}.
Coefficient estimates (\(\pm\) 1 standard deviation) are compared across
methods. This allows us to determine whether or not the main results
would differ depending on the method used.

@@

The data used in this manuscript is already publicly available (Pomeranz
et al. 2019b) data dryad DOI
\url{https://datadryad.org/stash/dataset/doi:10.5061/dryad.v6g985s},
(National Ecological Observatory Network (NEON) 2022). R Scripts to
reproduce the full simulation and analysis are available at:
({[}\url{https://github.com/Jpomz/detecting-spectra-differences}{]}).
(To be archived upon acceptance.)

@@

There was considerable variation in the \(\lambda\) estimate across
methodologies (Figure 1). The distribution of estimates from the MLE
method was always symmetrical and centered at the known value of
\(\lambda\) (Figure 1). The distribution of estimates from the binning
methods were generally wider and occasionally asymmetrical or bimodal.
On average, the 95\% confidence intervals produced by the NAS and ELBn
methods were \textasciitilde2 times wider than those produced by MLE
(Table 1), indicating greater consistency of estimates from MLE.
Similarly, estimates of \(\lambda\) deviated from the true \(\lambda\)
by an average of 0.04 or 0.05 absolute units for the NAS and ELBn
methods, more than 4 times higher than deviation (0.01) for the MLE
(Table 1).

\begin{figure}
\centering
\includegraphics{figures/est_lambda_est_b_density.png}
\caption{Figure 1. Distribution of Lambda estimates by method (color)
from random samples of body sizes from bounded power law distributions
with varying exponents (-2.5, -0.5). The figure is facetted by the known
lambda parameter (facet title) and is also shown as the dashed line in
each facet. Note that the x-axis varies in each facet.}
\end{figure}

\begin{longtable}[]{@{}llrrrr@{}}
\caption{Table 1. Summary of three methods in recapturing the known
values of lambda or the regression slopes simulated in this study.
Performance is determined by comparinged by uncertainty (range of 95\%
CI's) and the absolute distance of the model estimates from the known
estimates (median and sd of the difference). Values are summarized
across all n = 9000 or 6000 simulated datasets. See figures for more
specific comparisons.}\tabularnewline
\toprule
\begin{minipage}[b]{(\columnwidth - 5\tabcolsep) * \real{0.17}}\raggedright
Target\strut
\end{minipage} &
\begin{minipage}[b]{(\columnwidth - 5\tabcolsep) * \real{0.07}}\raggedright
Method\strut
\end{minipage} &
\begin{minipage}[b]{(\columnwidth - 5\tabcolsep) * \real{0.05}}\raggedleft
n\strut
\end{minipage} &
\begin{minipage}[b]{(\columnwidth - 5\tabcolsep) * \real{0.23}}\raggedleft
Median Range of 95\% CI\strut
\end{minipage} &
\begin{minipage}[b]{(\columnwidth - 5\tabcolsep) * \real{0.26}}\raggedleft
Median Absolute Deviation\strut
\end{minipage} &
\begin{minipage}[b]{(\columnwidth - 5\tabcolsep) * \real{0.22}}\raggedleft
SD Absolute Deviation\strut
\end{minipage}\tabularnewline
\midrule
\endfirsthead
\toprule
\begin{minipage}[b]{(\columnwidth - 5\tabcolsep) * \real{0.17}}\raggedright
Target\strut
\end{minipage} &
\begin{minipage}[b]{(\columnwidth - 5\tabcolsep) * \real{0.07}}\raggedright
Method\strut
\end{minipage} &
\begin{minipage}[b]{(\columnwidth - 5\tabcolsep) * \real{0.05}}\raggedleft
n\strut
\end{minipage} &
\begin{minipage}[b]{(\columnwidth - 5\tabcolsep) * \real{0.23}}\raggedleft
Median Range of 95\% CI\strut
\end{minipage} &
\begin{minipage}[b]{(\columnwidth - 5\tabcolsep) * \real{0.26}}\raggedleft
Median Absolute Deviation\strut
\end{minipage} &
\begin{minipage}[b]{(\columnwidth - 5\tabcolsep) * \real{0.22}}\raggedleft
SD Absolute Deviation\strut
\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.17}}\raggedright
Lambda\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.07}}\raggedright
MLE\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.05}}\raggedleft
9000\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.23}}\raggedleft
0.0661400\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.26}}\raggedleft
0.0120454\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.22}}\raggedleft
0.0191500\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.17}}\raggedright
Lambda\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.07}}\raggedright
NAS\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.05}}\raggedleft
9000\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.23}}\raggedleft
0.1335267\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.26}}\raggedleft
0.0460093\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.22}}\raggedleft
0.0929513\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.17}}\raggedright
Lambda\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.07}}\raggedright
ELBn\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.05}}\raggedleft
8579\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.23}}\raggedleft
0.1672598\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.26}}\raggedleft
0.0350186\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.22}}\raggedleft
0.0675122\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.17}}\raggedright
Regression Slope\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.07}}\raggedright
MLE\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.05}}\raggedleft
6000\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.23}}\raggedleft
0.0382185\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.26}}\raggedleft
0.0086982\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.22}}\raggedleft
0.0089766\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.17}}\raggedright
Regression Slope\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.07}}\raggedright
NAS\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.05}}\raggedleft
6000\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.23}}\raggedleft
0.0799201\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.26}}\raggedleft
0.0513434\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.22}}\raggedleft
0.0442020\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.17}}\raggedright
Regression Slope\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.07}}\raggedright
ELBn\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.05}}\raggedleft
6000\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.23}}\raggedleft
0.0923495\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.26}}\raggedleft
0.0300186\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 5\tabcolsep) * \real{0.22}}\raggedleft
0.0353218\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

@@

@@

All methods recaptured the correct sign of the slopes, yielding
qualitative consistency (Figure 4). However, the binning methods again
systematically underestimated the true value of the slope by
\textasciitilde0.05 units (Figure 5). Likewise, uncertainty in the slope
estimates was always greater in the binning methods, with the width of
the distributions increases with stronger relationships across a
hypothetical gradient. By comparison, the MLE showed no evidence of bias
and was always centered at the known value with relatively narrow
variation.

@@

The relationship between body size and abundance has been extensively
studied in a wide range of taxa inhabiting both terrestrial and aquatic
ecosystems (reviewed by {[}Brown (1995); GastonBlackburn2000; White et
al. (2007){]}). Empirical data shows generally consistent patterns and
can be explained by the metabolic theory of ecology (Brown et al. 2004).
Measuring parameters describing the decline in abundance with increasing
body size in communities is being done with increasing frequency across
ecology. Previous work has investigated the accuracy and inherent biases
associated with different estimation methods (White et al. 2007, Edwards
et al. 2017, Edwards et al. 2020). However, how these inaccuracies and
biases compound across environmental gradients remains uncertain, making
it difficult to detect variation in size spectra parameters across
environmental gradients with confidence. The most important outcome of
our work is that binning methods not only generate biased \(\lambda\)
values for individual data sets, but that bias carries over to affect
the parameters of subsequent regressions that use those \(\lambda\)'s as
response variables. This makes it challenging to understand how
\(\lambda\) varies in response to environmental gradients if binning is
used to estimate size spectra exponents.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
  \usepackage{amssymb}
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Manuscript: Detecting differences in Size Spectra},
  pdfauthor={Justin Pomeranz1,; James R. Junker2,3; Vojsava Gjoni4; Jeff S. Wesner4},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{lineno}
\usepackage{amsmath}
\usepackage{indentfirst}
\linenumbers
\newcommand{\beginsupplement}{ \setcounter{table}{0} \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\title{Manuscript: Detecting differences in Size Spectra}
\author{Justin Pomeranz\textsuperscript{1,*} \and James R.
Junker\textsuperscript{2,3} \and Vojsava
Gjoni\textsuperscript{4} \and Jeff S. Wesner\textsuperscript{4}}
\date{20 March, 2023}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The distribution of body size in communities is remarkably consistent
  across habitats and taxa and can be represented by size spectra, which
  are described by a power law. The focus of size spectra analysis is to
  estimate the exponent (\(\lambda\)) of the power law.
\item
  Many methods have been proposed for estimating \(\lambda\) most of
  which involve binning the data, summing abundance within bins, and
  then fitting a ordinary least squares (OLS) regression in log-log
  space. However, recent work has shown that binning procedures may
  return biased estimates of size spectra exponents compared to
  procedures that directly estimate \(\lambda\) using maximum likelihood
  estimation (MLE). Despite this variability in estimates, it is unclear
  if the relative change across environmental gradients is consistent
  across methodologies. Here, we used simulation to compare the ability
  of two binning methods (NAS, ELBn) and MLE to 1) recapture known
  values of \(\lambda\), and 2) recapture parameters in a linear
  regression measuring the change in \(\lambda\) across a hypothetical
  environmental gradient. We also compared the methods using two
  previously published body size datasets across a pollution gradient
  and a temperature gradient
\item
  Maximum likelihood methods always performed better than common binning
  methods, which demonstrated consistent bias depending on the simulated
  values of \(\lambda\). This bias carried over to the regressions,
  which were more accurate when \(\lambda\) was estimated using MLE
  compared to the binning procedures. Additionally, the variance in
  estimates using MLE methods is markedly reduced when compared to
  binning methods.
\item
  The uncertainty and variation in estimates when using binning methods
  can be of similar magnitudes as the variation previously published in
  experimental and observational studies, bringing into question the
  effect size of previously published results. However, while the
  methods produced different slope estimates from previously published
  datasets, they were in qualitative agreement on the sign of those
  slopes (i.e., all negative or all positive). Our results provide
  further support for the direct estimation of \(\lambda\) using MLE (or
  similar procedures) over the more common methods of binning.
\end{enumerate}

@@

We had two primary objectives in this study: 1) to compare how well
different methods estimate site-specific \(\lambda\)'s and 2) to compare
how different methods of estimating \(\lambda\) impact subsequent
regressions that use \(\lambda\)'s as a response variable. We did this
using both simulated and empirical data sets.

@@

After simulating data, we used three different methods (described below)
to estimate the value of \(\lambda\) (maximum likelihood, equal
logarithmic bins, and normalized abundance spectra). We repeated the
process of data simulation and parameter estimation 1000 times (reps)
and plotted the distribution of values obtained for each method.

@@

\hypertarget{equal-logarithmic-bins-elbn}{%
\subsection{Equal Logarithmic Bins:
ELBn}\label{equal-logarithmic-bins-elbn}}

In addition, to estimate the OLS slope parameter in log-log space, we
used two common binning methods. It is important to note that the
log-log regression is estimating \(\lambda\) as the regressionslope
(\(\beta_{1}\)), and not \(\lambda\) directly. In the interest of
clarity, we will refer to all estimates of lambda as \(\lambda\)
regardless of the method used, and reserve the terms ``parameters,''
``intercept,'' and ``slope'' \(\beta_{env}\) to refer to the parameters
of linear regressions that use the lambdas as a response variable.

For the first binning method, we created six equal logarithmic bins
covering the range of body sizes. This method has been used extensively
in previous studies (e.g., see Perkins et al. (2018), Martinez et al.
(2016), and Dossena et al. (2012). In the present study, the count in
each bin was normalized by dividing by the bin width to account for the
unequal bin sizes. The process of normalizing shifts the estimate by -1.
In other words, an un-normalized estimate of -0.75 would result in an
estimate of -1.75 when normalizing the data (Sprules and Barth 2015,
Edwards et al. 2017, Pomeranz et al. 2022). Throughout the manuscript,
the normalized equal logarithmic binning method will be referred to as
ELBn.

@@

The second binning method was similar to ELBn, but bins of
log\textsubscript{2} width are used. The number of bins is dependent on
the range of body sizes present in the data. When working with empirical
data with different size ranges this can alter the number of bins per
site, and it is recommended to construct the bins based on the global
minimum and maximum body size present. However, since the data here were
simulated from a known size range, the number of bins for each site is
identical. The count in each bin is normalized in the same way as
described above for the ELBn approach. This method has been used
extensively in the literature (e.g., Jennings et al. (2002, 2004),
Sprules and Barth (2015), Pomeranz et al. (2019a), McGarvey and Kirk
(2018)), and is referred to as the Normalized Abundance Spectrum (NAS).

@@

Where \(N\) is the count in each bin, \(M\) is the mid-point of each
bin, and \(\beta_{1}\) is considered to be the \(\lambda\) estimate.

@@

Previous work has investigated the biases in \(\lambda\) estimates when
using different methods (White et al. 2007, Edwards et al. 2017, Edwards
et al. 2020). However, the focus of the present work is to investigate
biases when estimating the relationship parameters across an
environmental gradient.

@@

\[\lambda_{ik} = \beta_{0} + \beta_{1}x_{ik}\]

@@

Once the body size data were simulated, we then reversed the procedure.
First, we used the three methods (NAS, ELBn, MLE) to estimate
\(\lambda\) for each of the 5000 body size data sets. Then, we fit a
linear regression to those data with the predictor variable \(x\) and
compared the resulting intercept (\(\beta_{0}\)) and slope
(\(\beta_{1}\)) parameter estimates to the known value. The main results
presented here were not dependent on the range of x-values or the number
of sites (Supplemental Information).

@@

The second data set was from the wadeable stream sites of NEON (National
Ecological Observatory Network (NEON) 2022). NEON stream sites are
located across a wide temperature gradient in the United States, from
Puerto Rico to Alaska. Quantitative macroinvertebrate samples were
collected using the most appropriate method based on the local habitat.
All individuals were identified and had their body lengths measured, and
body mass was estimated using published length-weight regressions. This
data has been analyzed previously using size spectra methods as
described in Pomeranz et al. (2022). Detailed methods of the sampling
collection and data processing methods can be found in the
macroinvertebrate data product information documents found on the NEON
website \url{https://data.neonscience.org/data-products/DP1.20120.001}.
Estimates of the slope coefficient (\(\beta_1\) ) \(\pm\) 1 SD, are
compared across methods. This allows us to determine whether or not the
main results would differ depending on the method used.

@@

The empirical data used in this manuscript is already publicly available
(Pomeranz et al. (2019b); data dryad DOI
\url{https://datadryad.org/stash/dataset/doi:10.5061/dryad.v6g985s}, and
(National Ecological Observatory Network (NEON) 2022). R Scripts to
reproduce the full simulation and analysis are available at:
({[}\url{https://github.com/Jpomz/detecting-spectra-differences}{]}).
(To be archived upon acceptance.)

@@

There was considerable variation in the \(\lambda\) estimate across
methodologies (Figure 1). The distribution of estimates from the MLE
method was always symmetrical and centered at the known value of
\(\lambda\) (Figure 1). The distribution of estimates from the binning
methods were generally wider and occasionally asymmetrical or bimodal.
On average, the 95\% confidence intervals produced by the NAS and ELBn
methods were \textasciitilde2 times wider than those produced by MLE
(Table 1), indicating greater consistency of estimates from MLE.
Similarly, estimates of \(\lambda\) deviated from the true \(\lambda\)
by an average of 0.04 or 0.05 absolute units for the NAS and ELBn
methods, more than four times higher than the deviation (0.01) observed
for the MLE (Table 1).

\begin{figure}
\centering
\includegraphics{figures/est_lambda_est_b_density.png}
\caption{Figure 1. Distribution of Lambda estimates by method (color)
from random samples of body sizes from bounded power law distributions
with varying exponents (-2.5 to -0.5). The figure is facetted by the
known lambda parameter (facet title) and is also shown as the dashed
line in each facet. Note that the x-axis varies in each facet.}
\end{figure}

@@

\hypertarget{table-1}{%
\paragraph{\texorpdfstring{\textbf{Table 1}}{Table 1}}\label{table-1}}

(\emph{Submitted as additional file on biorxiv.}) Summary of three
methods in recapturing the known values of lambda or the regression
slopes simulated in this study. Performance is determined by comparinged
by uncertainty (range of 95\% CI's) and the absolute distance of the
model estimates from the known estimates (median and SD of the
difference). Values are summarized across all n = 9000 or 6000 simulated
data sets. See figures for more specific comparisons.

@@

All methods recaptured the correct sign of the slopes, yielding
qualitative consistency (Figure 4). However, the binning methods again
systematically underestimated the true value of the slope by
\textasciitilde0.05 units (Figure 5). Likewise, uncertainty in the slope
estimates was always greater in the binning methods, with the width of
the distributions increasing with stronger relationships across a
hypothetical gradient. By comparison, the MLE showed no evidence of bias
and was always centered at the known value with relatively narrow
variation.

@@

The relationship between body size and abundance has been extensively
studied in a wide range of taxa inhabiting both terrestrial and aquatic
ecosystems (reviewed by Brown (1995), and White et al. (2007)).
Empirical data shows generally consistent patterns and can be explained
by the metabolic theory of ecology (Brown et al. 2004). Measuring
parameters describing the decline in abundance with increasing body size
in communities is being done with increasing frequency across ecology.
Previous work has investigated the accuracy and inherent biases
associated with different estimation methods (White et al. 2007, Edwards
et al. 2017, Edwards et al. 2020). However, the extent to which these
inaccuracies and biases compound across environmental gradients remains
uncertain, making it difficult to detect variation in size spectra
parameters across environmental gradients with confidence. The most
important outcome of our work is that binning methods not only generate
biased \(\lambda\) values for individual data sets, but that bias
carries over to affect the parameters of subsequent regressions that use
those \(\lambda\)'s as response variables. This makes it challenging to
understand how \(\lambda\) varies in response to environmental gradients
if binning is used to estimate size spectra exponents.
